Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-09 paper/paper.tex
Edits to Section 3

\subsection{Elements of Category Theory}
A \emph{category} is a collection of objects $A, B, C$, denoted by $\Obj(\CatC)$
and a collection of arrows $f, g, h$ between these objects, denoted by
$\Arr{\CatC}$, such that there always exists an identity arrow
$id_{A} : A \to A$ for each object $A$ and for two arrows $A \xrightarrow{f} B$
and $B \xrightarrow{g} C$ there always exists an arrow
$A \xrightarrow{g\circ f} C$ obeying the associativity law. We denote
$\Hom_{\CatC} (A, B)$ the set of arrows from $A$ to $B$ and we use the letters
$\CatC$, $\CatD$, $\CatE\dots$ for categories\footnote[1]{For presentation
purposes we shall not deal with size issues and assume all the categories are
locally small.}. The initial object in a category, denoted by $0$, is the object such
that for any other object $A$ there is a unique arrow $0 \xrightarrow{!} A$.
Dually, the terminal object, denoted by $1$, is the object such that for any
other object $A$ there is a unique arrow $A \xrightarrow{!} 1$. Initial and
terminal object are unique up-to isomorphism.

@@

In $\Set$, we can define the set of lists
\[
  \List{A} \iso 1 + A \times \List{A}
\]
as the set inductively generated by the
constructors $\nil : 1 \to \List{A}$ and
$\cons : A \times \List{A} \to \List{A}$.
The terminal object in $\Set$ is the \emph{unit} type, that only has one
inhabitant, $* : 1$.
The $\List{-}$ type is a functor. Its action on objects is to take any set
$A$ to $\List{A}$, and its functorial
action $\List{f} : \List{A} \to \List{B}$ is given by $\List{f}(*) = *$ and
$\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$. Notice that the definition
$\List{f}$ is well-defined as it recursively calls on a smaller argument.
However, there is a more general way of proving these type of definitions are
well-defined. Similarly, the set of streams $\Str{A} \iso A \times \Str{A}$ is
the greatest set generated by the constructor
$\cons : A \times \Str{A} \to \Str{A}$.

\subsection{Algebras and Catamorphisms}
\label{sec:algebras}
For a functor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(X,a_{X})$
where $X$ is an object of the category called the \emph{carrier} of the algebra
and $a_{X}$ is an arrow of type $FX \to X$ called the \emph{structure map}.
The category of $F$-algebras, denoted by $F\Alg(\CatC)$, is the category where
objects are $F$-algebras and arrows $f : (X, a_{X}) \to (Y, a_{Y})$ are
\emph{$F$-algebra homomorphisms}  $f : X \to Y$ in $\CatC$ such that they
respect the structure of the algebra, that is
$f \circ a_{X} = A_{Y} \circ F(f)$. The initial object in this category is
called the \emph{initial $F$-algebra}, that is the $F$-algebra which has a
unique $F$-algebra homomorphism into any other $F$-algebra. By Lambek's lemma
the initial $F$-algebra is precisely the \emph{least fixed-point} for the
functor $F$ which we denote by $(\mu F, \InIso)$ with
$\InIso : F \mu F \to \mu F$ and $\InOp : \mu F \to F \mu F$ witnessing the
isomorphism $F\mu F \iso \mu F$.  By the property of initial $F$-algebras, for
any other $F$-algebra $(X, a_{X})$ there exists a unique $F$-algebra
homomorphism, denoted by $\cata{\beta} : (\mu F, \InIso) \to (X, a_{X})$ and
pronounced ``\emph{catamorphism}'' satisfying:
\begin{equation}
  \label{eq:cata:unique}
  f = \cata{a_{X}} \iff f = a_{X} \circ F(f) \circ \InIso
\end{equation}
In words, any $F$-algebra homomorphism from the initial $F$-algebra is a
catamorphism. As a result of the uniqueness property we can derive the fusion
law. For all $F$-algebra homomorphisms $f : (X, a_{X}) \to (Y, a_{Y})$ we have
\begin{equation}
  \label{eq:cata:fusion}
  f \circ \cata{a_{X}} = \cata{a_{Y}}
\end{equation}
which means that the composition of a program $f$ with a catamorphism recursing
once over the data structure is the same as performing that recursion once using
the algebra $a_{Y}$ instead of $a_{X}$. This is a useful result for program
optimisation as we shall see. \mpav{We shall see? }

For example, for a set $A$ we define the functor $F : \Set \to \Set$ mapping
$X \mapsto 1 + A \times X$. An $F$-algebra is a set $B$ together with a
structure map $[b, i] : 1 + A \times B \to B$. The initial $F$-algebra is
clearly the set of lists $\List{A}$. The catamorphism associated with the type
of lists is the unique arrow recursively translating the initial algebra
$[\nil,\cons]$ into the algebra $[b, i]$.

\mpav{TODO}In functional programming this is commonly referred to as
$\foldr : (1 \to B) \to (A \times B \to B) \to B$. We can in fact set
$\foldr\;base\;ind = \cata{[base, ind]}$.

\subsection{Coalgebras and Anamorphisms}
\label{sec:coalg}
The dual of an algebra is a coalgebra. For an endofunctor $B : \CatC \to \CatC$,
a $B$-\emph{coalgebra} is a pair $(X, c_{X})$ where $X \in \Obj(\CatC)$ is the
carrier of the coalgebra and $c_{X} : X \to BX$ is a morphism.

For example, for a set of states $X$ and a finite set of labels $L$ we can
define a labelled transition system (LTS) on $X$ as a function $X \to BX$
implementing the \emph{transition system} with $BX = L \times X$.  In
particular, for a state $x_{1} \in X$, $c(x_{1})$ returns a pair $(l, x_{2})$
where $l \in L$ is the observable action and $x_{2} \in X$ is the next state.
In this case, we can even set some notation for the transition map:
\[
  x \xrightarrow{l} y \eqdef c(x) = (l, y)
\]
The category of $B$-coalgebras, denoted $B\Coalg$, is the category where objects
are $B$-coalgebras and morphisms $f : (X, \gamma_{X}) \to (Y, \gamma_{Y})$ are
$B$-coalgebra homomorphisms $f : A \to B$, that is
$ \gamma_{Y} \circ f = F(f) \circ \gamma_{X}$. Terminal object in this category
$B$-coalgebra corresponds to the greatest fixed-point for the functor $B$,
denoted by $(\nu B, \OutIso)$ where $\OutIso$ denotes the final $B$-coalgebra
$\OutIso : \nu B \to B\nu B$.

For example, the terminal coalgebra for the functor $BX = L \times X$ is the set
of of infinite streams, that is the greatest solution to the domain equation
$\Str{A} \iso A \times \Str{A}$.  As a result, for any $B$-coalgebra
$(B, \gamma)$ there exists a unique $B$-coalgebra homomorphism into the terminal
coalgebra $(\nu B, \OutIso)$ which is denoted by $\ana{\gamma}$ and pronounced
``\emph{anamorphism}''. We spell out the uniqueness property:
\begin{equation}
  \label{eq:ana:unique}
  f = \ana{\gamma} \iff f = \OutOp \circ B(f) \circ \gamma
\end{equation}
In words, for any $B$-coalgebra, if there is any other $B$-coalgebra
homomorphism $f$ into the terminal object then it must be the anamorphism on the
same coalgebra. As a result of the uniqueness property we can derive the fusion
laws. For all $B$-coalgebra homomorphisms
$f : (X, \gamma_{X}) \to (Y, \gamma_{Y})$ we have
\begin{equation}
  \label{eq:ana:fusion}
  \ana{\gamma_{Y}} \circ f = \ana{\gamma_{X}}
\end{equation}

@@

Let $(A,a)$ be an $F$-algebra and $(C,c)$ be an $F$-coalgebra. An arrow
$C \to A$ is an \emph{hylomorphism}, written $h : (C,c) \rightarrowtail (A,a)$ if
it satisfies
\begin{equation}
  \label{eq:hylo}
  h = a \circ F(h) \circ c
\end{equation}
As we stated earlier, a solution to this equation does not exist for an
arbitrary algebra and coalgebra pair and, in fact, this definition cannot be
accepted by Coq.

A coalgebra $(C,c)$ is \emph{recursive} if for every algebra $(A, a)$ there is a
\emph{unique} hylo $(C,c ) \rightarrowtail (A, a)$. We denote these type of
hylos by $\hylo{c}{a}$.

An example of a recursive coalgebra is the \coq{partition} function in quicksort
which destructures a list into a pivot and two sublist and as long as the
sublists are smaller the partitioning function still yields a unique solution to
the recursion scheme.
The uniqueness property of the hylo yields the following fusion law:
\begin{equation}
  \label{eq:hylo:fusion}
  f \circ \hylo{c}{a}= \hylo{c'}{a}
\end{equation}
Using the hylo fusion law, we can prove the well-known \emph{deforestation
optimisation}. This is when two consecutive recursive computations, one that
builds a data structure, and another one that consumes it, can be fused together
into a single recursive definition. This, in turn, allows us to prove that a
recursive hylomorphism is the composition of a catamorphism and a recursive
anamorphism.

$(A, c_{1})$ is an recursive coalgebra
$(B, c_{2})$ is an recursive coalgebra

$(C, a_{2})$ is an algebra
$(B, a_{1})$ is an algebra

$c_{2} \circ a_{1} = id$

\begin{equation}
  \label{eq:hylo:deforestation}
  \hylo{a_{2}}{c_{2}} \circ \hylo{a_{1}}{c_{1}} = \hylo{a_{2}}{c_{1}}
\end{equation}

\begin{coqcode}
Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
  (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
  : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.
\end{coqcode}

\subsection{Recursive Anamorphisms}
Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
inductive data type in the following way. A recursive coalgebra
can be applied only finitely many times, therefore when this is applied to an
anamorphism the only possible traces it can produce from the seed function are
the finite ones. We denote this special kind of recursion scheme \emph{recursive
anamorphism}. We can show that recursive anamorphisms of type $X \to \nu F$
can also be given the type $X \to \mu F$. Moreover, these anamorphisms are
exactly hylomorphisms on the recursive $F$-coalgebra and the
algebra $\InIso$ for the inductive data type $\nu F$. This fact falls out
from the universality property of the hylomorphism and the fact that recursive
anamorphisms satisfy the same equation.

@@

\begin{minted}{haskell}
data ListF a x = Nil | Cons a x deriving Functor
data TreeF a x = Empty | Node x a x deriving Functor
\end{minted}

@@

In this section, we focus on mechanising \emph{container functors} in a way that
is suitable for code extraction. Recall from Section~\ref{sec:rec-schemes} that
we need to be able toabstract away from the particular shape of the data, and
that this is achieved by introducing the concept of \emph{functors} which have
the suitable fixed-point properties, i.e.\ those which have a initial algebras.
A common approach to construct such functors is to use
\emph{containers}~\cite{AbbottAG05}.  However, reasoning about container
equality will require us to consider both functional extensionality and
heterogeneous equality. We avoid these axioms by introducing a custom
equivalence relation on types (Section~\ref{sec:types-morphisms}).

@@

\paragraph{Containers} Another common solution is to the strict positivity
problem is the use \emph{containers}~\cite{AbbottAG05}.  A container
$S \triangleright P$ is defined by a type of \emph{shapes} $S : \Type$ and a
\emph{family of position types} indexed by shapes $P : S \to \Type$.  An
\emph{extension} of this container is a functor
$\llbracket S \triangleright P \rrbracket$, whose action on objects is given by
$\llbracket S \triangleright P \rrbracket \; X = \Sigma s : S.\; P\;s \to X$,
and on morphisms is given by post-composition, keeping the shape the same
$\llbracket S \triangleright P \rrbracket \; f = \lambda (s,\;g).\;(s,\;f\circ g)$.

To explain the role of shapes and positions in containers, consider the functor
$F\; X = 1 + A \times X$.  An equivalent container, i.e.\ a container with an
extension that is isomorphic to $F$ requires a shape that can distinguish two
cases, indicating that there are two constructors to build an object of type
$F\;X$. For example, we could define $S_{F} = 1 + A$. We need to consider two
cases for the family of positions:  $P_{F}(\mathsf{inj}_{1}\;*) = 0$, because
there are no occurrences of $X$ in the left case; and
$P_{F}(\mathsf{inj}_{2}\;a) = 1$, because there is one occurrence of $X$ in the
right hand side. It is easy to see that
$\llbracket S_{F} \triangleright P_{F} \rrbracket \cong F$.

In general, every polynomial functor has an equivalent container, with an
isomorphic extension. We have mechanised this result, and this can  be found in
the accompanying code (file Container.v).
% 
%Generally, any functor that
%is isomorphic to a container extension is called a \emph{container functor}.
%Here we will call \emph{container functor} the functors defined by a specific
%container extension.

@@

\subsubsection{Equality of Container Extensions}
Reasoning about the equality of container extensions is not entirely solved by
using decidable validity predicates to define the families of positions. In
general, we want to equate container extensions that have the same shape, and
that, for equal shape and position, they return the same element. To avoid the
use of the functional extensionality axiom, we capture this relation with the
following inductive proposition in Coq:
\begin{coqcode}
Inductive AppR (x y : App F X) : Prop :=
| AppR_ext (shape x = shape y)
           (forall p1 p2, projT1 e1 = projT1 e2 -> contents p1 = contents p2).
\end{coqcode}
Note that  we do not care about the validity proof of the positions, only their
value. This is to simplify (slightly) our proofs. This relation is
trivially \emph{reflexive}, \emph{transitive}, and \emph{symmetric}.

However, the use of a different equality for container extensions now forces us
to deal with the fact that some types have different definitions of equality.
In particular, we want to reason about the equality of functions of types such
as \coq{App A -> B} (or \coq{B -> App A}). Since
these types now come with their own equivalence, any function that manipulates
them needs to be \emph{respectful}. I.e.\ given 
\coq{R : X -> X -> Prop} and \coq{R' : Y -> Y -> Prop},
we want functions (morphisms) that satisfy the following property:
\begin{coqcode}
forall (x y : X), R x y -> R' (f x) (f y)
\end{coqcode}

@@

We prove the isomorphisms of polynomial types and the equivalent 
container extensions. As an example, we discuss (informally) the isomorphisms
of pairs with their equivalent container extensions. Suppose that we
know that \coq{App F X} is isomorphic to \coq{A}, and
\coq{App G X} is isomorphic to \coq{B}. Then we can
show that \coq{App (Prod F G) X} is isomorphic to
\coq{A * B}.  If we have an element of type 
\coq{App (Prod F G) X},
using the \coq{inl} position, we can obtain
\coq{App F X}.  Similarly, using \coq{inr}, we can
obtain \coq{App G X}. Since these are the only two valid positions
in the shape of pairs, we have finished. It is now sufficient to use the
isomorphisms of \coq{App F X} and \coq{App G X} to
obtain \coq{A * B}. Similarly if we have \coq{A * B},
we can first use the isomorphisms of 
\coq{A} and
\coq{B} to obtain
\coq{App F X * App G X}, and then construct the necessary container
extension. Given \coq{p_inl : Pos l -> Pos (l * r)} (resp. \coq{p_inr}) that
act as \coq{inl} (resp.\ \coq{inr}) on product positions, and
\coq{case_pos : (Pos l -> X) -> (Pos r -> X) -> Pos (l * r) -> X} that pattern
matches on the product positions, the functions that witness the isomorphism
are:
\begin{coqcode}
Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
  ({| shape := shape (fst x); cont := fun e => cont x (p_inl e) |}, 
   {| shape := shape (snd x); cont := fun e => cont x (p_inr e) |}).
Definition iso_prod (x : App F X * App G X) : App (Prod F G) X :=
{| shape := (shape (fst x), shape (snd x));
   cont := case_pos (cont (fst x)) (cont (snd x)) |}.
\end{coqcode}
Proving that the composition of these functions is the identity is
straightforward using the fact that \coq{Prod} containers only have two valid
positions.

>>>>

\subsection{Elements of Category Theory}
A \emph{category} is a collection of objects $A, B, C$, denoted by $\Obj(\CatC)$
and a collection of arrows $f, g, h$ between these objects, denoted by
$\Arr{\CatC}$, such that there always exists an identity arrow
$id_{A} : A \to A$ for each object $A$ and for two arrows $A \xrightarrow{f} B$
and $B \xrightarrow{g} C$ there always exists an arrow
$A \xrightarrow{g\circ f} C$ obeying the associativity law. We denote
$\Hom_{\CatC} (A, B)$ the set of arrows from $A$ to $B$ and we use the letters
$\CatC$, $\CatD$, $\CatE\dots$ for categories\footnote[1]{For presentation
purposes we shall not deal with size issues and assume all the categories are
small.}. The initial object in a category, denoted by $0$, is the object such
that for any other object $A$ there is a unique arrow $0 \xrightarrow{!} A$.
Dually, the terminal object, denoted by $1$, is the object such that for any
other object $A$ there is a unique arrow $A \xrightarrow{!} 1$. Initial and
terminal object are unique up-to isomorphism.

@@

In $\Set$, we can define the set of lists
\[
  \List{A} \iso 1 + A \times \List{A}
\]
as the set inductively generated by the
constructors $\nil : 1 \to \List{A}$ and
$\cons : A \times \List{A} \to \List{A}$.
The terminal object in $\Set$ is the \emph{unit} type, that only has one
inhabitant, $* : 1$, analogous to the unit type in Haskell: \haskell{() :: ()}.
The $\List{-}$ type is a functor. Its action on objects is to take any set
$A$ to $\List{A}$, and its functorial
action $\List{f} : \List{A} \to \List{B}$ is given by $\List{f}(*) = *$ and
$\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$. Notice that the definition
$\List{f}$ is well-defined as it recursively calls on a smaller argument.
However, there is a more general way of proving these type of definitions are
well-defined. Similarly, the set of streams $\Str{A} \iso A \times \Str{A}$ is
the greatest set generated by the constructor
$\cons : A \times \Str{A} \to \Str{A}$.

\subsection{Algebras and Catamorphisms}
\label{sec:algebras}
For a functor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(A,\alpha)$
where $A$ is an object of the category called the \emph{carrier} of the algebra
and $\alpha$ is an arrow of type $FA \to A$ called the \emph{structure map}.
The category of $F$-algebras, denoted by $F\Alg(\CatC)$, is the category where
objects are $F$-algebras and arrows $f : (A, \alpha) \to (B, \beta)$ are
\emph{$F$-algebra homomorphisms}  $f : A \to B$ in $\CatC$ such that they
respect the structure of the algebra, that is
$f \circ \alpha = \beta \circ F(f)$. The initial object in this category is
called the \emph{initial $F$-algebra}, that is the $F$-algebra which has a
unique $F$-algebra homomorphism into any other $F$-algebra. By Lambek's lemma
the initial $F$-algebra is precisely the \emph{least fixed-point} for the
functor $F$ which we denote by $(\mu F, \InIso)$ with
$\InIso : F \mu F \to \mu F$ and $\InOp : \mu F \to F \mu F$ witnessing the
isomorphism $F\mu F \iso \mu F$.
By the property of initial $F$-algebras, for
any other $F$-algebra $(B, \beta)$ there exists a unique $F$-algebra
homomorphism, denoted by $\cata{\beta} : (\mu F, \InIso) \to (B, \beta)$ and
pronounced ``\emph{catamorphism}'' satisfying:
\begin{equation}
  \label{eq:cata:unique}
  f = \cata{\alpha} \iff f = \alpha \circ F(f) \circ \InIso
\end{equation}
In words, any $F$-algebra homomorphism from the initial $F$-algebra is a
catamorphism.

For example, for a set $A$ we define the functor $F : \Set \to \Set$ mapping
$X \mapsto 1 + A \times X$. An $F$-algebra is a set $B$ together with a
structure map $[base, ind] : 1 + A \times B \to B$. The initial $F$-algebra is
clearly the set of lists $\List{A}$. The catamorphism associated with the type
of lists is the unique arrow recursively translating the initial algebra
$[\nil,\cons]$ into the algebra $[base, ind]$:

In functional programming this is commonly referred to as
$\foldr : (1 \to B) \to (A \times B \to B) \to B$. We can in fact set
$\foldr\;base\;ind = \cata{[base, ind]}$.

\subsection{Coalgebras and Anamorphisms}
\label{sec:coalg}
The dual of an algebra is a coalgebra. For an endofunctor $B : \CatC \to \CatC$,
a $B$-\emph{coalgebra} is a pair $(X, c)$ where $X \in \Obj(\CatC)$ is the
carrier of the coalgebra and $c : X \to BX$ is a morphism.

For example, for a set of states $X$ and a finite set of labels $L$ we can
define a labelled transition system (LTS) on $X$ as a function $X \to BX$
implementing the \emph{transition system} with $BX = L \times X$.  In
particular, for a state $x \in X$, $c(x_{1})$ returns a pair $(l, x_{2})$ where
$l \in L$ is the observable action and $x_{2} \in X$ is the next state.  In this
case, we can even set some notation for the transition map:
\[
  x \xrightarrow{l} y \eqdef c(x) = (l, y)
\]
The category of $B$-coalgebras, denoted $B\Coalg$, is the category where objects
are $B$-coalgebras and morphisms $f : (X, \gamma_{X}) \to (Y, \gamma_{Y})$ are
$B$-coalgebra homomorphisms $f : A \to B$, that is
$ \gamma_{Y} \circ f = F(f) \circ \gamma_{X}$. Terminal object in this category
$B$-coalgebra corresponds to the greatest fixed-point for the functor $B$,
denoted by $(\nu B, \OutIso)$ where $\OutIso$ denotes the final $B$-coalgebra
$\OutIso : \nu B \to B\nu B$. For

For example, the terminal coalgebra for the functor $BX = L \times X$ is the set
of of infinite streams, that is the greatest solution to the domain equation
$\Str{A} \iso A \times \Str{A}$.  As a result, for any $B$-coalgebra
$(B, \gamma)$ there exists a unique $B$-coalgebra homomorphism into the terminal
coalgebra $(\nu B, \OutIso)$ which is denoted by $\ana{\gamma}$ and pronounced
\emph{anamorphism}. We spell out the uniqueness property:
\begin{equation}
  \label{eq:ana:unique}
  f = \ana{\gamma} \iff f = \OutOp \circ B(f) \circ \gamma
\end{equation}
In words, for any $B$-coalgebra, if there is any other $B$-coalgebra
homomorphism $f$ into the terminal object then it must be the anamorphism on the
same coalgebra.

@@

In order to find a solution we restrict ourselves to the so-called
\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.
A $F$-coalgebra $(X, c)$ is recursive if

An example of a
recursive coalgebra is the \coq{partition} function in quicksort which
destructures a list into a pivot and two sublist and as long as the sublists are
smaller the partitioning function still yields a unique solution to the
recursion scheme.

\begin{coqcode}
Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
  (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.
\end{coqcode}
Using the hylo fusion law, we can prove the well-known \emph{deforestation
optimisation}. This is when two consecutive recursive computations, one that
builds a data structure, and another one that consumes it, can be fused together
into a single recursive definition. This, in turn, allows us to prove that a
recursive hylomorphism is the composition of a catamorphism and a recursive
anamorphism.
\begin{coqcode}
Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
  (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
  : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.
\end{coqcode}

\subsection{Recursive Anamorphisms}
Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
inductive data type in the following way.  Intuitively, a recursive coalgebra
can be applied only finitely many times, therefore when this is applied to an
anamorphism the only possible traces it can produce from the seed function are
the finite ones. We denote this special kind of recursion scheme \emph{recursive
anamorphism}.  We can show that recursive anamorphisms of type \coq{X -> GFix F}
can also be given the type \coq{X -> LFix F}. Moreover, these anamorphisms are
exactly hylomorphisms on the recursive \coq{F}-coalgebra and the
algebra \coq{In} for the inductive data type \coq{LFix F}. This fact falls out
from the universality property of the hylomorphism and the fact that recursive
anamorphisms satisfy the same equation.

@@

@@

In this section, we focus on mechanising \emph{container functors} in a way that
is suitable for code extraction. Recall from Section~\ref{sec:rec-schemes} that
we need to be able to abstract away from particular shape of the data, and we
can do this using \emph{functors} that have suitable fixed-point properties;
i.e.\ those which have a initial algebras.
%
A common approach to construct such functors is to use
\emph{containers}~\cite{AbbottAG05}. However, reasoning about container equality
will require us to consider both functional extensionality and heterogeneous
equality. We avoid these axioms by introducing a custom equivalence relation on
types (Section~\ref{sec:types-morphisms}).

@@

% \paragraph{Containers}
Another common solution is to the strict positivity
problem is the use \emph{containers}~\cite{AbbottAG05}.  A container
$S \triangleright P$ is defined by a type of \emph{shapes} $S : \Type$ and a
\emph{family of position types} indexed by shapes $P : S \to \Type$.  An
\emph{extension} of this container is a functor
$\llbracket S \triangleright P \rrbracket$, whose action on objects is given by
$\llbracket S \triangleright P \rrbracket \; X = \Sigma s : S.\; P\;s \to X$,
and on morphisms is given by post-composition, keeping the shape the same
$\llbracket S \triangleright P \rrbracket \; f = \lambda (s,\;g).\;(s,\;f\circ g)$.

To explain the role of shapes and positions in containers, consider the functor
$F\; X = 1 + A \times X$.  An equivalent container, i.e.\ a container with an
extension that is isomorphic to $F$, requires a shape that can distinguish two
cases, indicating that there are two constructors to build an object of type
$F\;X$. For example, we could define $S_{F} = 1 + A$. We need to consider two
cases for the family of positions:  $P_{F}(\mathsf{inj}_{1}\;*) = 0$, because
there are no occurrences of $X$ in the left case; and
$P_{F}(\mathsf{inj}_{2}\;a) = 1$, because there is one occurrence of $X$ in the
right hand side. It is easy to see that
$\llbracket S_{F} \triangleright P_{F} \rrbracket \cong F$.

In general, every polynomial functor has an equivalent container, with an
isomorphic extension, and we have mechanised this result.
% and this can  be found in the accompanying code (file Container.v).
% 
%Generally, any functor that
%is isomorphic to a container extension is called a \emph{container functor}.
%Here we will call \emph{container functor} the functors defined by a specific
%container extension.

@@

\subsubsection{Equality of Container Extensions}
Reasoning about the equality of container extensions is not entirely solved by
using decidable validity predicates to define the families of positions. In
general, we want to equate container extensions that have the same shape, and
that, for equal shape and position, they return the same element. To avoid the
use of the functional extensionality axiom, we capture this relation with the
following inductive proposition:
\begin{coqcode}
Inductive AppR (x y : App F X) : Prop :=
| AppR_ext (shape x = shape y)
           (forall p1 p2, projT1 e1 = projT1 e2 -> contents p1 = contents p2).
\end{coqcode}
Note that we do not care about the validity proof of the positions, only their
value. This is to simplify (slightly) our proofs. This relation is trivially
\emph{reflexive}, \emph{transitive}, and \emph{symmetric}.

However, the use of a different equality for container extensions now forces us
to deal with the fact that some types have different definitions of equality.
In particular, we want to reason about the equality of functions of types such
as \coq{App A -> B} (or \coq{B -> App A}).
%
Since these types now come with their own equivalence, any function that
manipulates them needs to be \emph{respectful}; i.e.\ given
\coq{R : X -> X -> Prop} and \coq{R' : Y -> Y -> Prop},
we want functions (morphisms) that satisfy the following property:
\begin{coqcode}
forall (x y : X), R x y -> R' (f x) (f y)
\end{coqcode}

@@

We prove the isomorphisms of polynomial types and the equivalent 
container extensions. As an example, we will consider the isomorphisms
of pairs with their equivalent container extensions. Suppose that we
know that \coq{App F X} is isomorphic to \coq{A}, and
\coq{App G X} is isomorphic to \coq{B}. Then we can
show that \coq{App (Prod F G) X} is isomorphic to
\coq{A * B}.  If we have an element of type 
\coq{App (Prod F G) X},
using the \coq{inl} position, we can obtain
\coq{App F X}.  Similarly, using \coq{inr}, we can
obtain \coq{App G X}. Since these are the only two valid positions
in the shape of pairs, we have finished. It is now sufficient to use the
isomorphisms of \coq{App F X} and \coq{App G X} to
obtain \coq{A * B}. Similarly if we have \coq{A * B},
we can first use the isomorphisms of 
\coq{A} and
\coq{B} to obtain
\coq{App F X * App G X}, and then construct the necessary container
extension. Given \coq{p_inl : Pos l -> Pos (l * r)} (resp. \coq{p_inr}) that
act as \coq{inl} (resp.\ \coq{inr}) on product positions, and
\coq{case_pos : (Pos l -> X) -> (Pos r -> X) -> Pos (l * r) -> X} that pattern
matches on the product positions, the functions that witness the isomorphism
are:
\begin{coqcode}
Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
  ({| shape := shape (fst x); cont := fun e => cont x (p_inl e) |}, 
   {| shape := shape (snd x); cont := fun e => cont x (p_inr e) |}).
Definition iso_prod (x : App F X * App G X) : App (Prod F G) X :=
{| shape := (shape (fst x), shape (snd x));
   cont := case_pos (cont (fst x)) (cont (snd x)) |}.
\end{coqcode}
Proving that the composition of these functions is the identity is
straightforward using the fact that \coq{Prod} containers only have two valid
positions.
